Next Trippy Structures\thanks{Supported by organization x.}

### 2019-03-07 paper.tex
Bridge for XCTR

\subsection{\ctr}
\label{sec:ctr}
A fundamental weakness of \ttctr~is that it requires several binary search operations over the \texttt{CSA} in the following cases:
\begin{itemize}
    \item We are interested in the number of passengers that started their trip at a stop $s$ and a time window $t_a...t_z$, but from \textbf{any line} (\texttt{start\_X$_{T}$} from Section~\ref{sec:rq}). As $jcodes$ are relative to lines, we must make a separate query for each possible $(s:l_i) \forall l_i \in L$ combinations.
    \item We need to restrict the line of a final stop, in queries such as \texttt{end\_X$_{L}$} or \texttt{from\_X\_to\_Y$_{L}$} (and similar variations). Because the final stops belong to separate entries of the vocabulary that do not encode line identifiers, to restrict a stop $Y\in S$ to a line $l\in L$ we need to search for every possible expanded pattern $W:l,Y...$, for every stop $W$ from the line $l$ that could have been boarded before alighting at $Y$. While it is tempting to address this issue by modifying the design of \ttctr so that final stops also encode line identifiers, this would in turn make queries that do not restrict the line of the final stop inefficient, and we would need to perform a new query for every combination of $(s:l_i) \forall l_i \in L$, as in the previous case. \marginpar{\tiny Viendo los experimentos, uno se puede preguntar si no habria quedado mejor meter lineas en paradas finales...}
\end{itemize}

Bla bla we need the same sorted trips as in \ttctr. $V$ is not important anymore.

We use three complementary structures to represent each component of the sequence, as shown in Figure~\ref{fig:example_xctr}:
\begin{enumerate}[(i)]
    \item An adapted Compressed Suffix Array (\texttt{CSA}) over the stop identifiers of all trips, concatenated into a string with additional terminator symbols $\$$ appended at the end of each trip. While in the final CSA we assign these $\$$ a lexicographical value of 0 (thus smaller than any stop identifier), we assign them different values during the construction of the suffix array (A) to ensure that the entries for $\$$ in A maintain the same order as in the original text. Finally, we make a modification on $\Psi$ to make the entries of each $\$$ point to the start of its own trip instead of the next one. These two modifications are proven necessary for our implemented queries, at the expense of losing some of the properties of a classic CSA.
    \item \texttt{WML}: Aligned to the entries of (i) there is a Wavelet Matrix (WM) for the line identifiers of each stop. Aligned to the $\$$ section we duplicate the starting lines of each trip. As a trivial optimization, we build a separate WM for every stop, allowing us to save space due to the fact that a single stop does not usually belong to many lines, thus the average height of these WM is no larger (and usually smaller) than the height of a single WM.
    \item \texttt{WMJ}: A WM of jcodes aligned to the leaves of (ii). Note that this makes this structure dependant on (ii), which is coherent with the fact that journey codes themselves are relative to the line identifier. In case (ii) implements the optimization described, the entries of the WM must also be rearranged to match the delimited stops.
\end{enumerate}

@@

A more complex query that we can solve with \ctr~is ``number of trips that started from a stop \DataSty{s$_a$} and ended at a stop \DataSty{s$_z$}'', which can be further restricted to specific starting and ending lines and a time window. The pseudocode for the full version of such query can be found at the Algorithm~\ref{alg:xy}. This algorithm relies heavily on the abstract function \FuncSty{GetRange}, which when applied to a CSA, delimits the range of entries of $\Psi$ that match a queried string pattern, while on a WM it works as the $range$ operation defined in Section~\ref{sec:wm}, also reporting the range in the last level of a WM corresponding to the entries of a single queried symbol within a range in the encoded sequence (equivalent to two \FuncSty{TrackDown} operations on the first and last occurrence of the queried symbol, but more time efficient).

>>>>

\subsection{\ctr}
\label{sec:ctr}
A fundamental weakness of \ttctr~is that it requires several binary search operations over the \texttt{CSA} in the following cases:
\begin{itemize}
    \item We are interested in the number of passengers that started their trip at a stop $s$ and a time window $t_a...t_z$, but from \textbf{any line} (\texttt{start\_X$_{T}$} from Section~\ref{sec:rq}). As $jcodes$ are relative to lines, we must make a separate query for each possible $(s:l_i) \forall l_i \in L$ combinations.
    \item We need to restrict the line of a final stop, in queries such as \texttt{end\_X$_{L}$} or \texttt{from\_X\_to\_Y$_{L}$} (and similar variations). Because the final stops belong to separate entries of the vocabulary that do not encode line identifiers, to restrict a stop $Y\in S$ to a line $l\in L$ we need to search for every possible expanded pattern $W:l,Y...$, for every stop $W$ from the line $l$ that could have been boarded before alighting at $Y$. While it looks tempting to address this issue by modifying the design of \ttctr~so that final stops also encode line identifiers, this would in turn make queries that do not restrict the line of the final stop inefficient, and we would need to perform a new query for every combination of $(s:l_i) \forall l_i \in L$, as in the previous case. \marginpar{\tiny Viendo los experimentos, uno se puede preguntar si no habria quedado mejor meter lineas en paradas finales...}
\end{itemize}

These weakness motivated us to develop a second version, \ctr, which reduces the complexity of the queries that restrict the final line, and delegates line checks on a new WM, allowing for more space-time trade-off. As in \ttctr, the input trips need to be sorted by the same criteria, but in \ctr~we use three complementary structures to represent each component of the sequence, as shown in Figure~\ref{fig:example_xctr}:
\begin{enumerate}[(i)]
    \item An adapted Compressed Suffix Array (\texttt{CSA}) over the stop identifiers of all trips, concatenated into a string with additional terminator symbols $\$$ appended at the end of each trip. As in the \texttt{CSA} from \ttctr, we make these $\$$ symbols maintain the order of the trips and cyclical in $\Psi$. Because this time we do not encode line and stop identifiers together and \texttt{CSA} only encodes stops, there is no need for a complex vocabulary anymore.
    \item \texttt{WML}: Aligned to the entries of (i) there is a Wavelet Matrix (WM) for the line identifiers of each stop. Aligned to the $\$$ section we duplicate the starting lines of each trip. As a trivial optimization, we build a separate WM for every stop, allowing us to save space due to the fact that a single stop does not usually belong to many lines, thus the average height of these WM is no larger (and usually smaller) than the height of a single WM.
    \item \texttt{WMJ}: A WM of jcodes aligned to the leaves of (ii). Note that this makes this structure dependant on (ii), which is coherent with the fact that journey codes themselves are relative to the line identifier. In case (ii) implements the optimization described, the entries of the WM must also be rearranged to match the delimited stops.
\end{enumerate}

@@

A more complex query that we can solve with \ctr~is ``number of trips that started from a stop \DataSty{s$_a$} and ended at a stop \DataSty{s$_z$}'', which can be further restricted to specific starting and ending lines and a time window (\texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} in Section~\ref{sec:rq}). The pseudocode for the full version of such query can be found at the Algorithm~\ref{alg:xy}. This algorithm relies heavily on the abstract function \FuncSty{GetRange}, which when applied to a CSA, delimits the range of entries of $\Psi$ that match a queried string pattern, while on a WM it works as the $range$ operation defined in Section~\ref{sec:wm}, also reporting the range in the last level of a WM corresponding to the entries of a single queried symbol within a range in the encoded sequence (equivalent to two \FuncSty{TrackDown} operations on the first and last occurrence of the queried symbol, but more time efficient).
